% APPEARANCE.TEX
%
% The documentation in this file is part of PyXPlot
% <http://www.pyxplot.org.uk>
%
% Copyright (C) 2006-7 Dominic Ford <coders@pyxplot.org.uk>
%               2008   Ross Church
%
% $Id$
%
% PyXPlot is free software; you can redistribute it and/or modify it under the
% terms of the GNU General Public License as published by the Free Software
% Foundation; either version 2 of the License, or (at your option) any later
% version.
%
% You should have received a copy of the GNU General Public License along with
% PyXPlot; if not, write to the Free Software Foundation, Inc., 51 Franklin
% Street, Fifth Floor, Boston, MA  02110-1301, USA

% ----------------------------------------------------------------------------

% LaTeX source for the PyXPlot Users' Guide

\chapter{Labelling Plots and Producing Galleries}

\section{Adding Arrows, Text Labels, and More}

This section describes how to put arrows and text labels on plots; the syntax
is similar to that used by gnuplot, but slightly changed. It is now possible,
for example, to set the linestyles and colours with which arrows should be
drawn.  Also covered is how to put grids onto plots, and how to change the size
and colour of textual labels on plots.

\subsection{Arrows}

\label{set_arrow}\index{arrows} Arrows may be placed on plots using the
\indcmdt{set arrow}, which has similar syntax to that used by gnuplot. A simple
example would be:

\begin{verbatim}
set arrow 1 from 0,0 to 1,1
\end{verbatim}

\noindent The number `1' immediately following `set arrow' specifies an identification
number for the arrow, allowing it to be subsequently removed via:

\begin{verbatim}
unset arrow 1
\end{verbatim}

\noindent or equivalently, via:\indcmd{set noarrow}

\begin{verbatim}
set noarrow 1
\end{verbatim}

In PyXPlot, this syntax is extended; the {\tt set arrow} command can be
followed by the keyword `{\tt with}', to specify the style of the arrow. For
example, the specifiers `{\tt nohead}', `{\tt head}' and
`{\tt twohead}', after the keyword `{\tt with}', can be used to make
arrows with no arrow heads, normal arrow heads, or two arrow heads.
`{\tt twoway}' is an alias for `{\tt twohead}'.  For example:

\begin{verbatim}
set arrow 1 from 0,0 to 1,1 with nohead
\end{verbatim}

In addition, linestyles and colours can be specified after the keyword
`{\tt with}':

\begin{verbatim}
set arrow 1 from 0,0 to 1,1 with nohead \
linetype 1 c blue
\end{verbatim}

As in gnuplot, the coordinates for the start and end points of the arrow can be
specified in a range of coordinate systems. `{\tt first}', the default,
measures the graph using the $x$- and $y$-axes. `{\tt second}' uses the $x2$-
and $y2$-axes. `{\tt screen}' and `{\tt graph}' both measure in centimetres
from the origin of the graph. In the following example, we use these
specifiers, and specify coordinates using variables rather than doing so
explicitly:

\begin{verbatim}
x0 = 0.0
y0 = 0.0
x1 = 1.0
y1 = 1.0
set arrow 1 from first  x0, first  x1 \
            to   screen x1, screen x1 \
            with nohead
\end{verbatim}

In addition to these four options, which are those available in gnuplot, the
syntax `{\tt axis}\textit{n}' may also be used, to use the $n\,$th $x$- or
$y$-axis -- for example, `{\tt axis3}'.\indcmd{set arrow} This allows arrows to
reference any arbitrary axis on plots which make use of large numbers of
parallel axes (see Section~\ref{multiple_axes}).

\subsection{Text Labels}

Text labels may be placed on plots using the \indcmdt{set label}. As with all
textual labels in PyXPlot, these are rendered in \LaTeX:

\begin{verbatim}
set label 1 'Hello World' at 0,0
\end{verbatim}

As in the previous section, the number `1' is a reference number, which allows
the label to be removed by either of the following two commands:

\begin{verbatim}
set nolabel 1
unset label 1
\end{verbatim}

\noindent The positional coordinates for the text label, placed after the
keyword `{\tt at}', can be specified in any of the coordinate systems
described for arrows above. A rotation angle may optionally be specified after
the keyword `{\tt rotate}', to rotate text counter-clockwise by a given
angle, measured in degrees. For example, the following would produce
upward-running text:

\begin{verbatim}
set label 1 'Hello World' at axis3 3.0, axis4 2.7 rotate 90
\end{verbatim}

\index{fontsize}\index{text!size} The fontsize of these text labels can
globally be set using the \indcmdt{set fontsize}. This applies not only to the
{\tt set label} command, but also to plot titles, axis labels, keys, etc. The
value given should be an integer in the range $-4 \leq x \leq 5$. The default
is zero, which corresponds to \LaTeX's {\tt normalsize}; $-4$ corresponds to
{\tt tiny} and 5 to {\tt Huge}.

\index{text!colour}\index{colours!text} The \indcmdt{set textcolour} can be
used to globally set the colour of all text output, and applies to all of the
text that the {\tt set fontsize} command does. It is especially useful when
producing plots to be embedded in presentation slideshows, where bright text on
a dark background may be desired. It should be followed either by an integer,
to set a colour from the present palette, or by a colour name. A list of the
recognised colour names can be found in Section~\ref{colour_names}.  For
example:

\begin{verbatim}
set textcolour 2
set textcolour blue
\end{verbatim}

\index{text!alignment}\index{alignment!text}By default, each label's specified
position corresponds to its bottom left corner. This alignment may be changed
with the {\tt set texthalign}\indcmd{set texthalign} and {\tt set
textvalign}\indcmd{set textvalign} commands. The former takes the options {\tt
left}, {\tt centre} or {\tt right}, and the latter takes the options {\tt
bottom}, {\tt centre} or {\tt top}, for example:

\begin{verbatim}
set texthalign right
set textvalign top
\end{verbatim}

\subsection{Gridlines}

Gridlines may be placed on a plot and subsequently removed via the statements:

\begin{verbatim}
set grid
set nogrid
\end{verbatim}

\noindent respectively. The following commands are also valid:

\begin{verbatim}
unset grid
unset nogrid
\end{verbatim}

\noindent By default, gridlines are drawn from the major and minor ticks of the
$x$- and $y$-axes. However, the axes which should be used may be specified
after the \indcmdt{set grid}\index{grid}:

\begin{verbatim}
set grid x2y2
set grid x x2y2
\end{verbatim}

\noindent The top example would connect the gridlines to the ticks of the $x2$-
and $y2$- axes, whilst the lower would draw gridlines from both the $x$- and
the $x2$-axes.

If one of the specified axes does not exist, then no gridlines will be drawn in
that direction.  Gridlines can subsequently be removed selectively from some
axes via:

\begin{verbatim}
unset grid x2x3
\end{verbatim}

The colours of gridlines\index{grid!colour}\index{colours!grid} can be
controlled via the {\tt set gridmajcolour}\indcmd{set gridmajcolour} and {\tt
set gridmincolour}\indcmd{set gridmincolour} commands, which control the
gridlines emanating from major and minor axis ticks respectively. An example
would be:

\begin{verbatim}
set gridmincolour blue
\end{verbatim}

\noindent Any of the colour names listed in Section~\ref{colour_names} can be
used.

A related command\index{axes!colour}\index{colours!axes} is {\tt set
axescolour}\indcmd{set axescolour}, which has a syntax similar to that above,
and sets the colour of the graph's axes.\label{set_colours}

\section{LaTeX and PyXPlot}

\section{Multi-plotting}
\label{multiplot}
\index{multiplot}

Gnuplot has a plotting mode called `multiplot' which allows many graphs to be
plotted together, and display side-by-side. The basic syntax of this mode is
reproduced in PyXPlot, but is hugely extended.

The mode is entered by the command {\tt set multiplot}\indcmd{set multiplot}.
This can be compared to taking a blank sheet of paper on which to place plots.
Plots are then placed on that sheet of paper, as usual, with the {\tt plot}
command. The position of each plot is set using the \indcmdt{set origin}, which
takes a comma-separated $x,y$ coordinate pair, measured in centimetres. The
following, for example, would plot a graph of $\sin(x)$ to the left of a plot
of $\cos(x)$:

\begin{verbatim} 
set multiplot
plot sin(x)
set origin 10,0
plot cos(x)
\end{verbatim}

The multiplot page may subsequently be cleared with the \indcmdt{clear}, and
multiplot mode may be left using the \indcmdt{set nomultiplot}.

\subsection{Deleting, Moving and Changing Plots}

Each time a plot is placed on the multiplot page in PyXPlot, it is allocated a
reference number, which is output to the terminal. Reference numbers count up
from zero each time the multiplot page is cleared. A number of commands exist
for modifying plots after they have been placed on the page, selecting them by
making reference to their reference numbers.

Plots may be removed from the page with the \indcmdt{delete}, and restored with
the \indcmdt{undelete}:

\begin{verbatim} 
delete <number>
undelete <number>
\end{verbatim}

The reference numbers of deleted plots are not reused until the page is
cleared, as they may always be restored with the \indcmdt{undelete}; plots
which have been deleted simply do not appear.

Plots may also be moved with the \indcmdt{move}. For example, the following
would move plot 23 to position (8,8) measured in centimetres:

\begin{verbatim} 
move 23 to 8,8
\end{verbatim}

In multiplot mode, the \indcmdt{replot} can be used to modify the last plot
added to the page. For example, the following would change the title of the
latest plot to `foo', and add a plot of $\cos(x)$ to it:

\begin{verbatim} 
set title 'foo'
replot cos(x)
\end{verbatim}

Additionally, it is possible to modify any plot on the page, by first selecting
it with the \indcmdt{edit}. Subsequently, the \indcmdt{replot} will act upon
the selected plot. The following example would produce two plots, and then
change the colour of the text on the first:

\begin{verbatim} 
set multiplot
plot sin(x)
set origin 10,0
plot cos(x)
edit 0        # Select the first plot ...
set textcolour red
replot        # ... and replot it.
\end{verbatim}

The \indcmdt{edit} can also be used to view the settings which are applied to
any plot on the multiplot page -- after executing {\tt edit~0}, the
\indcmdt{show} will show the settings applied to plot zero.

When a new plot is added to the page, {\tt replot} always switches to act
upon this most recent plot.

\index{replotting} The \indcmdt{refresh} is rather similar to the
\indcmdt{replot}, but produces an exact copy of the latest display. This can be
useful, for example, after changing the terminal type, to produce a second copy
of a multiplot page in a different format. But the crucial difference between
this command and {\tt replot} is that it doesn't replot anything. Indeed, there
could be only textual items and arrows on the present multiplot page, and no
graphs \textit{to} replot.

\subsection{Linked Axes}

The axes of plots can be linked together, in such a way that they always share
a common scale. This can be useful when placing plots next to one another,
firstly, of course, if it is of intrinsic interest to ensure that they are on a
common scale, but also because the two plots then do not both need their own
axis labels, and space can be saved by one sharing the labels from the other.
In PyXPlot, an axis which borrows its scale and labels from another is called a
`linked axis'.

Such axes are declared by setting the label of the linked axis to a magic
string such as {\tt linkaxis 0}\label{linked_axes}\index{axes!reserved
labels}\index{magic axis labels}. This magic label would set the axis to borrow
its scale from an axis from plot zero. The general syntax is `{\tt linkaxis}
$n$ $m$', where $n$ and $m$ are two integers, separated by a comma or
whitespace. The first, $n$, indicates the plot from which to borrow an axis;
the second, $m$, indicates whether to borrow the scale of axis $x1$, $x2$,
$x3$, etc. By default, $m=1$. The linking will fail, and a warning result, if
an attempt is made to link to an axis which doesn't exist.

\subsection{Text Labels, Arrows and Images}

\label{text_command} In addition to placing plots on the multiplot page, text
labels may also be inserted independently of any plots, using the
\indcmdt{text}. This has the following syntax:

\begin{verbatim} 
text 'This is some text' at x,y
\end{verbatim}

In this case, the string `This is some text' would be rendered at position
$(x,y)$ on the multiplot. As with the \indcmdt{set label}, a rotation angle may
optionally be specified to rotate text labels through any given angle, measured
in degrees counter-clockwise, for example:

\begin{verbatim} 
text 'This is some text' at x,y rotate r
\end{verbatim}

The commands {\tt set textcolour}, {\tt set
texthalign} and {\tt set textvalign}, which have already been described in
the context in the {\tt set label} command, can also be used to set the
colour and alignment of text produced with the {\tt text} command.\index{set
textcolour command@{\tt set textcolour} command}\index{set texthalign
command@{\tt set texthalign} command}\index{set textvalign
command@{\tt set textvalign} command}. A useful application of this is to
produce centred headings at the top of multiplots.

As with plots, each text item has a unique identification number, and can be
moved around, deleted or undeleted with the {\tt delete}, {\tt undelete}
and {\tt move} commands.
\index{delete command@{\tt delete} command}
\index{undelete command@{\tt undelete} command}

It should be noted that the {\tt text} command can also be used outside of
the multiplot environment, to render a single piece of short text instead of a
graph. One obvious application is to produce equations rendered as graphical
files for inclusion in talks.

\label{arrows} \index{arrow command@{\tt arrow} command} Arrows may also be
placed on multiplot pages, independently of any plots, using the {\tt arrow}
command, which has syntax:

\begin{verbatim} 
arrow from x,y to x,y
\end{verbatim}

As above, arrows receive unique identification numbers, and can be deleted and
undeleted.

The {\tt arrow} command may be followed by the `{\tt with}' keyword to
specify to style of the arrow. The style keywords which are accepted are
identical to those accepted by the {\tt set arrow} command (see
Section~\ref{set_arrow}). For example:

\begin{verbatim} 
arrow from x1,y1 to x2,y2 \
with twohead colour red
\end{verbatim}

\index{jpeg command@{\tt jpeg} command} Bitmap images in jpeg form may be
placed on the multiplot using the {\tt jpeg} command.  This has syntax:

\begin{verbatim}
jpeg 'filename' at x,y width w
\end{verbatim}

As an alternative to the {\tt width} modifier the height of the image can be
specified, using the analogous {\tt height} modifier.  An optional angle can
also be specified using the {\tt rotate} modifier; this causes the included
image to be rotated counter-clockwise by a specified angle (in degrees).

\index{eps command@{\tt eps} command} Vector graphic images in eps format may
be placed on to a multiplot using the {\tt eps} command, which has a syntax
analogous to the {\tt jpeg} command.  However neither height nor width need be
specified; in this case the image will be included at its native size.  For
example:

\begin{verbatim}
eps 'filename' at 3,2 rotate 5
\end{verbatim}

\noindent will place the eps file with its bottom-left corner at position
$(3,2)$ cm from the origin, rotated counter-clockwise through 5 degrees.

\subsection{Speed Issues}
\label{set_display}

By default, whenever an item is added to a multiplot, or an existing item moved
or replotted, the whole multiplot is replotted to show the change. This can be
a time consuming process on large and complex multiplots. For this reason, the
{\tt set nodisplay}\index{set display command@{\tt set display} command}
command is provided, which stops PyXPlot from producing any output. The
{\tt set display} command can subsequently be issued to return to normal
behaviour.

This can be especially useful in scripts which produce large multiplots. There
is no point in producing output at each step in the construction of a large
multiplot, and so a great speed increase can be achieved by wrapping the script
with:

\begin{verbatim} 
set nodisplay
[...prepare large multiplot...]
set display
refresh
\end{verbatim}

