% MISC.TEX
%
% The documentation in this file is part of PyXPlot
% <http://www.pyxplot.org.uk>
%
% Copyright (C) 2006-7 Dominic Ford <coders@pyxplot.org.uk>
%               2008   Ross Church
%
% $Id$
%
% PyXPlot is free software; you can redistribute it and/or modify it under the
% terms of the GNU General Public License as published by the Free Software
% Foundation; either version 2 of the License, or (at your option) any later
% version.
%
% You should have received a copy of the GNU General Public License along with
% PyXPlot; if not, write to the Free Software Foundation, Inc., 51 Franklin
% Street, Fifth Floor, Boston, MA  02110-1301, USA

% ----------------------------------------------------------------------------

% LaTeX source for the PyXPlot Users' Guide

\chapter{PyXPlot and the Outside World}

This Chapter describes a few additional features of PyXPlot.

\section{The Commandline Environment}

PyXPlot uses the GNU Readline command-line environment, which means that the up
and down arrow keys can be used to repeat previously executed commands. Each
user's command history is stored in his homespace in a history file called
`\texttt{.pyxplot\_history}', allowing PyXPlot to remember command histories
between sessions. Additionally, a \texttt{save} command is provided, allowing
the user to save his command history from the present session to a text file;
this has the following syntax:

\begin{verbatim}
save 'output_filename'
\end{verbatim}

The related \index{history command@\texttt{history} command} prints the history
to the terminal.

From the shell command line, the PyXPlot accepts the following switches which
modify its behaviour:\index{command line syntax}

\begin{longtable}{p{3.5cm}p{8.5cm}}
\texttt{-h --help} & Display a short help message listing the available command-line switches.\\
\texttt{-v --version} & Display the current version number of PyXPlot.\\
\texttt{-q --quiet} & Turn off the display of the welcome message on startup. \\
\texttt{-V --verbose} & Display the welcome message on startup, as happens by default. \\
\texttt{-c --colour} & Use colour highlighting\footnote{This will only function on terminals which support colour output.} to display output in green, warning messages in amber, and error messages in red.\footnote{The author apologies to those members of the population who are red/green colourblind, but draws their attention to the following sentence.} These colours can be changed in the \texttt{terminal} section of the configuration file; see Section~\ref{configuration} for more details. \\
\texttt{-m --monochrome} & Do not use colour highlighting, as happens by default. \\
\end{longtable}

\section{Variables}

PyXPlot recognises two types of variables; numeric variables and string
variables.  The former can be assigned a value using any valid mathematical
expression, for example

\begin{verbatim}
a = 5.2 * sqrt(64)
\end{verbatim}

would assign the value 41.6 to the variable ``a''.  Numerical variables can
subsequently be used in mathematical expressions themselves, for example:

\begin{verbatim}
a=2*pi
plot [0:1] sin(x) with lines
\end{verbatim}

String variables can be assigned to in an analogous manner by enclosing the
string in quotation marks.  They can then be used wherever a string would be
required, for example

\begin{verbatim}
plotname = "A very exciting graph"
set title plotname
\end{verbatim}

String variables can also be modified using the search-and-replace string
operator, =~.  This takes a regular expression in standard form and applies it
to the relvent string.  For example:

\begin{verbatim}
twister="seven silver soda syphons"
twister =~ s/s/th/
print twister
\end{verbatim}

\section{Formatting and Terminals}
\label{set_terminal2}

In this section I shall outline the new and modified commands for controlling
the graphic output format of PyXPlot.

The widths of plots may be set be means of two commands -- \texttt{set
size}\index{set size command@\texttt{set size} command} and \texttt{set
width}\index{set width command@\texttt{set width} command}. Both are
equivalent, and should be followed by the desired width measured in
centimetres, for example:

\begin{verbatim}
set width 20
\end{verbatim}

The \texttt{set size} command can also be used to set the aspect ratio of plots
by following it with the keyword \texttt{ratio}\index{set size ratio
command@\texttt{set size ratio} command}. The number which follows should be
the desired ratio of height to width. The following, for example would produce
plots three times as high as they are wide:

\begin{verbatim}
set size ratio 3.0
\end{verbatim}

The command \texttt{set size noratio} returns to PyXPlot's default aspect ratio
of the golden ratio, i.e. $\left((1+\sqrt{5})/2\right)^{-1}$, which matches
that of a sheet of A4 paper\footnote{Of less practical significance, it has
been in use since the time of the Pythagoreans, and is seen repeatedly in the
architecture of the Parthenon.}.  The special command \texttt{set size
square}\index{set size square command@\texttt{set size square} command} sets
the aspect ratio to unity.

If the \texttt{enlarge} modifier is used with the {\tt set terminal} command
then the whole plot is enlarged or, in the case of large plots, shrunk to the
current paper size (minus a small margin).  The aspect ratio of the plot is
preserved.  This is perhaps most useful when preparing a plot to send to a
printer with the postscript terminal.

In Section~\ref{directing_output} I described how the \texttt{set terminal}
command\index{set terminal command@\texttt{set terminal} command} can be used
to produce plots in various graphic formats. In addition, I here describe how
the way in which plots are displayed on the screen can be changed. The default
terminal, \texttt{X11}, is used to send output to screen.

By default, each time a new plot is generated, if the previous plot is still
open on the display, the X11 terminal will replace it with the new one, thus
keeping only one plot window open at a time. This has the advantage that the
desktop does not become flooded with plot windows.

If this behaviour is not desired, old plots can be kept visible when plotting
further graphs by using the the \texttt{X11\_multiwindow} terminal: \index{X11
terminal}\index{multiple windows}

\begin{verbatim} 
set terminal X11_singlewindow
plot sin(x)
plot cos(x)  <-- first plot window disappears
\end{verbatim}

\noindent c.f.:

\begin{verbatim} 
set terminal X11_multiwindow
plot sin(x)
plot cos(x)  <-- first plot window remains
\end{verbatim}

As an additional option, the \texttt{X11\_persist} terminal keeps plot windows
open after PyXPlot exits; the above two terminals close all plot windows upon
exit.

\index{set terminal command@\texttt{set terminal} command} As there are many
changes to the options accepted by the \texttt{set terminal} command in
comparison to those understood by gnuplot, the syntax of PyXPlot's command is
given below, followed by a list of the recognised settings:

\begin{verbatim} 
set terminal { X11_singlewindow | X11_multiwindow | X11_persist |
               postscript | eps | pdf | gif | png | jpg }
             { colour | color | monochrome }
             { portrait | landscape }
             { invert | noinvert }
             { transparent | solid }
             { enlarge | noenlarge }
\end{verbatim}

\begin{longtable}{p{3cm}p{9cm}}
\texttt{x11\_singlewindow} & Displays plots on the screen (in X11 windows, using ghostview). Each time a new plot is generated, it replaces the old one, preventing the desktop from becoming flooded with old plots.\footnote{The author is aware of a bug, that this terminal can occasionally go blank when a new plot is generated. This is a known bug in ghostview, and can be worked around by selecting File $\to$ Reload within the ghostview window.} \textbf{[default when running interactively; see below]}\\
\end{longtable} % WHY IS THIS NECESSARY??
\begin{longtable}{p{3cm}p{9cm}}
\texttt{x11\_multiwindow} & As above, but each new plot appears in a new window, and the old plots remain visible. As many plots as may be desired can be left on the desktop simultaneously.\\
\texttt{x11\_persist} & As above, but plot windows remain open after PyXPlot closes.\\
\texttt{postscript} & Sends output to a postscript file. The filename for this file should be set using \texttt{set output}. \textbf{[default when running non-interactively; see below]}\index{postscript output}\\
\texttt{eps} & As above, but produces encapsulated postscript.\index{encapsulated postscript}\index{postscript!encapsulated}\\
\texttt{pdf} & As above, but produces pdf output.\index{pdf output}\\
\texttt{gif} & Sends output to a gif image file; as above, the filename should be set using \texttt{set output}.\index{gif output}\\
\texttt{png} & As above, but produces a png image.\index{png output}\\
\texttt{jpg} & As above, but produces a jpeg image.\index{jpeg output}\\
\texttt{colour} & Allows datasets to be plotted in colour. Automatically they will be displayed in a series of different colours, or alternatively colours may be specified using the \texttt{with colour} plot modifier (see below). \textbf{[default]}\index{colour output}\\
\texttt{color} & Equivalent to the above; provided for users of nationalities which can't spell. \smiley \\
\texttt{monochrome} & Opposite to the above; all datasets will be plotted in black.\index{monochrome output}\\
\texttt{portrait} & Sets plots to be displayed in upright (normal) orientation. \textbf{[default]}\index{portrait orientation}\\
\texttt{landscape} & Opposite of the above; produces side-ways plots. Not very useful when displayed on the screen, but you fit more on a sheet of paper that way around.\index{landscape orientation}\\
\texttt{invert} & Modifier for the gif, png and jpg terminals; produces output with inverted colours.\footnote{This terminal setting is useful for producing plots to embed in talk slideshows, which often contain bright text on a dark background. It only works when producing bitmapped output, though a similar effect can be achieved in postscript using the \texttt{set textcolour} and \texttt{set axescolour} commands (see Section~\ref{set_colours}).}\index{colours!inverting}\\
\texttt{noinvert} & Modifier for the gif, png and jpg terminals; opposite to the above. \textbf{[default]}\\
\texttt{transparent} & Modifier for the gif and png terminals; produces output with a transparent background.\index{transparent terminal}\index{gif output!transparency}\index{png output!transparency}\\
\texttt{solid} & Modifier for the gif and png terminals; opposite to the above. \textbf{[default]}\\
\texttt{enlarge} & Enlarge or shrink contents to fit the current paper
size.\index{enlarging output}\\
\texttt{noenlarge} & Do not enlarge output; opposite to the above. \textbf{[default]}\\
\end{longtable}
\label{terminals}

The default terminal is normally \texttt{x11\_singlewindow}, matching
approximately the behaviour of gnuplot. However, there is an exception to this.
When PyXPlot is used non-interactively -- i.e. one or more command scripts are
specified on the command line, and PyXPlot exits as soon as it finishes
executing them -- the \texttt{x11\_singlewindow} is not a very sensible
terminal to use. Any plot window would close as soon as PyXPlot exited. The
default terminal in this case changes to \texttt{postscript}.

One exception to this is when the special `--' filename is specified in a list
of command scripts on the command line, to produce an interactive terminal
between running a series of scripts. In this case, PyXPlot detects that the
session will be interactive, and defaults to the usual
\texttt{x11\_singlewindow} terminal.

An additional exception is on machines where the \texttt{DISPLAY} environment
variable\index{display environment variable@\texttt{DISPLAY} environment
variable} is not set. In this case, PyXPlot detects that it has access to no
X-terminal on which to display plots, and defaults to the \texttt{postscript}
terminal.

The \texttt{gif}, \texttt{png} and \texttt{jpg} terminals result in some loss
of quality, since the plot has to be sampled into a bitmapped graphic format.
By default, this sampling is performed at 300 dpi, though this may be changed
using the command \texttt{set dpi <value>}. Alternatively, it may be changed
using the \texttt{DPI} option in the \texttt{settings} section of a
configuration file (see Section~\ref{configuration}).\index{set dpi
command@\texttt{set dpi} command}\index{resolution of bitmap
output}\index{image resolution}

\section{Paper Sizes}

By default, when the postscript terminal produces printable, i.e. not
encapsulated, output, the paper size for this output is read from your system
locale settings. It may be changed, however, with the \texttt{set
papersize}\index{set papersize command@\texttt{set papersize} command} command,
which may be followed either by the name of a recognised paper size, or by the
dimensions of a user-defined size, specified as a \texttt{height},
\texttt{width} pair, both being measured in millimetres. For example:

\begin{verbatim}
set papersize a4
set papersize 100,100
\end{verbatim}

A list of recognised paper size names is given in Figure~\ref{paper_sizes}.

\begin{figure}
\tiny \center
\begin{tabular}{|rrr|rrr|}
\hline
\textbf{Name} & \textbf{$h$/mm} & \textbf{$w$/mm} & \textbf{Name} & \textbf{$h$/mm} & \textbf{$w$/mm} \\
\hline
                       2a0 &   1681 &   1189 &           medium &    584 &    457 \\
                       4a0 &   2378 &   1681 &          monarch &    267 &    184 \\
                        a0 &   1189 &    840 &             post &    489 &    394 \\
                        a1 &    840 &    594 &        quad\_demy &   1143 &   889 \\
                       a10 &     37 &     26 &           quarto &    254 &    203 \\
                        a2 &    594 &    420 &            royal &    635 &    508 \\
                        a3 &    420 &    297 &        statement &    216 &    140 \\
                        a4 &    297 &    210 &       swedish\_d0 &   1542 &   1090 \\
                        a5 &    210 &    148 &       swedish\_d1 &   1090 &    771 \\
                        a6 &    148 &    105 &      swedish\_d10 &     48 &     34 \\
                        a7 &    105 &     74 &       swedish\_d2 &    771 &    545 \\
                        a8 &     74 &     52 &       swedish\_d3 &    545 &    385 \\
                        a9 &     52 &     37 &       swedish\_d4 &    385 &    272 \\
                        b0 &   1414 &    999 &       swedish\_d5 &    272 &    192 \\
                        b1 &    999 &    707 &       swedish\_d6 &    192 &    136 \\
                       b10 &     44 &     31 &       swedish\_d7 &    136 &     96 \\
                        b2 &    707 &    499 &       swedish\_d8 &     96 &     68 \\
                        b3 &    499 &    353 &       swedish\_d9 &     68 &     48 \\
                        b4 &    353 &    249 &       swedish\_e0 &   1241 &    878 \\
                        b5 &    249 &    176 &       swedish\_e1 &    878 &    620 \\
                        b6 &    176 &    124 &      swedish\_e10 &     38 &     27 \\
                        b7 &    124 &     88 &       swedish\_e2 &    620 &    439 \\
                        b8 &     88 &     62 &       swedish\_e3 &    439 &    310 \\
                        b9 &     62 &     44 &       swedish\_e4 &    310 &    219 \\
                        c0 &   1296 &    917 &       swedish\_e5 &    219 &    155 \\
                        c1 &    917 &    648 &       swedish\_e6 &    155 &    109 \\
                       c10 &     40 &     28 &       swedish\_e7 &    109 &     77 \\
                        c2 &    648 &    458 &       swedish\_e8 &     77 &     54 \\
                        c3 &    458 &    324 &       swedish\_e9 &     54 &     38 \\
                        c4 &    324 &    229 &       swedish\_f0 &   1476 &   1044 \\
                        c5 &    229 &    162 &       swedish\_f1 &   1044 &    738 \\
                        c6 &    162 &    114 &      swedish\_f10 &     46 &     32 \\
                        c7 &    114 &     81 &       swedish\_f2 &    738 &    522 \\
                        c8 &     81 &     57 &       swedish\_f3 &    522 &    369 \\
                        c9 &     57 &     40 &       swedish\_f4 &    369 &    261 \\
                     crown &    508 &    381 &       swedish\_f5 &    261 &    184 \\
                      demy &    572 &    445 &       swedish\_f6 &    184 &    130 \\
               double\_demy &    889 &    597 &       swedish\_f7 &    130 &     92 \\
                  elephant &    711 &    584 &       swedish\_f8 &     92 &     65 \\
               envelope\_dl &    110 &    220 &       swedish\_f9 &     65 &     46 \\
                 executive &    267 &    184 &       swedish\_g0 &   1354 &    957 \\
                  foolscap &    330 &    203 &       swedish\_g1 &    957 &    677 \\
         government\_letter &    267 &    203 &      swedish\_g10 &     42 &     29 \\
international\_businesscard &     85 &     53 &       swedish\_g2 &    677 &    478 \\
               japanese\_b0 &   1435 &   1015 &       swedish\_g3 &    478 &    338 \\
               japanese\_b1 &   1015 &    717 &       swedish\_g4 &    338 &    239 \\
              japanese\_b10 &     44 &     31 &       swedish\_g5 &    239 &    169 \\
               japanese\_b2 &    717 &    507 &       swedish\_g6 &    169 &    119 \\
               japanese\_b3 &    507 &    358 &       swedish\_g7 &    119 &     84 \\
               japanese\_b4 &    358 &    253 &       swedish\_g8 &     84 &     59 \\
               japanese\_b5 &    253 &    179 &       swedish\_g9 &     59 &     42 \\
               japanese\_b6 &    179 &    126 &       swedish\_h0 &   1610 &   1138 \\
               japanese\_b7 &    126 &     89 &       swedish\_h1 &   1138 &    805 \\
               japanese\_b8 &     89 &     63 &      swedish\_h10 &     50 &     35 \\
               japanese\_b9 &     63 &     44 &       swedish\_h2 &    805 &    569 \\
            japanese\_kiku4 &    306 &    227 &       swedish\_h3 &    569 &    402 \\
            japanese\_kiku5 &    227 &    151 &       swedish\_h4 &    402 &    284 \\
         japanese\_shiroku4 &    379 &    264 &       swedish\_h5 &    284 &    201 \\
         japanese\_shiroku5 &    262 &    189 &       swedish\_h6 &    201 &    142 \\
         japanese\_shiroku6 &    188 &    127 &       swedish\_h7 &    142 &    100 \\
                large\_post &    533 &    419 &       swedish\_h8 &    100 &     71 \\
                    ledger &    432 &    279 &       swedish\_h9 &     71 &     50 \\
                     legal &    356 &    216 &          tabloid &    432 &    279 \\
                    letter &    279 &    216 &  us\_businesscard &     89 &     51 \\
\hline
\end{tabular}
\caption{A list of all of the named paper sizes recognised by the \texttt{set papersize}\index{set
papersize command@\texttt{set papersize} command} command, with their heights, $h$, and widths, $w$, measured in millimetres.}
\label{paper_sizes}
\end{figure}

\section{Script Watching: pyxplot\_watch}

PyXPlot includes a simple tool for watching command script files, and executing
them whenever they are modified. This may be useful when developing a command
script, if one wants to make small modifications to it, and see the results in
a semi-live fashion. This tool is invoked by calling the
\texttt{pyxplot\_watch}\index{pyxplot\_watch}\index{watching scripts} command
from a shell prompt. The command-line syntax of \texttt{pyxplot\_watch} is
similar to that of PyXPlot itself, for example:

\begin{verbatim}
pyxplot_watch script
\end{verbatim}

\noindent would set \texttt{pyxplot\_watch} to watch the command script file
\texttt{script}. One difference, however, is that if multiple script files are
specified on the command line, they are watched and executed independently,
\textit{not} sequentially, as PyXPlot itself would do. Wildcard characters can
also be used to set \texttt{pyxplot\_watch} to watch multiple
files.\footnote{Note that \texttt{pyxplot\_watch *.script} and
\texttt{pyxplot\_watch $\backslash$*.script} will behave differently in most
UNIX shells.  In the first case, the wildcard is expanded by your shell, and a
list of files passed to \texttt{pyxplot\_watch}. Any files matching the
wildcard, created after running \texttt{pyxplot\_watch}, will not be picked up.
In the latter case, the wildcard is expanded by \texttt{pyxplot\_watch} itself,
which \textit{will} pick up any newly created files.}

This is especially useful when combined with GhostView's watch facility. For
example, suppose that a script \texttt{foo} produces postscript output
\texttt{foo.ps}. The following two commands could be used to give a live view
of the result of executing this script:

\begin{verbatim}
gv --watch foo.ps &
pyxplot_watch foo
\end{verbatim}
